Follow-up Samples

### 2017-01-20 submission/manuscript_outline_20161024.Rmd
clean up text edits

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or before samples in order to confirm that the community was what was responding or changing due to surgery.  We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to comensal bacteria.  With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT both models (lesion and before sample) had a significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].  Further confirmation of the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occured.  First, one potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.  Second, the bacteria even if they were reduced in relative abundance they were still present within the gut.  Third, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Fourth, is that it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.  Finally, it is also possible that all three of these potential explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

@@

Another interesting outcome was that Within figure 3 the before sample model showed better test AUC results then the training set AUC.  This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be overfit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.

>>>>

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or before samples in order to confirm that the community was what was responding or changing due to surgery.  We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to comensal bacteria.  With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT both models (lesion and before sample) had a significantly decreased positive probability of either lesion or being a before sample on follow up [Figure 4 & S4].  Further confirmation of the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occured.  First, one potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.  Second, the bacteria even if they were reduced in relative abundance they were still present within the gut.  Third, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Fourth, is that it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.  Finally, it is also possible that all of these potential explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

@@

Another interesting outcome was that within figure 3 the before sample model showed better test AUC results then the training set AUC.  This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be overfit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.

### 2017-01-20 submission/manuscript_outline_20161024.tex
clean up text edits

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or before samples in order to confirm that the community was what was
responding or changing due to surgery. We found that the OTUs that made
up the most important variables to the model overwhelmingly belonged to
comensal bacteria. With only the lesion model having a single OTU from a
previously associated cancer bacterium (\emph{Porphyromonas
asaccharolytica}). Using only these important OTUs and FIT both models
(lesion and before sample) had a significantly decreased positive
probability of either lesion or being an initial sample on follow up
{[}Figure 4 \& S4{]}. Further confirmation of the importance of the
changes of comensal bacteria to these classifications was that a total
of 14 OTUs were common to both models and the vast majority belonged to
regular residents of our gut community.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occured. First, one potential explanation is that even with surgery
and a shift of the bacterial community these specific bacteria still
persisted within the colon. Second, the bacteria even if they were
reduced in relative abundance they were still present within the gut.
Third, is that they were not present in enough individuals to be able to
classify those with and without disease with a high degree of accuracy.
Fourth, is that it is possible that our Random Forest models were able
to get the same information from measures such as FIT or other OTUs.
Finally, it is also possible that all three of these potential
explanations could have played a role. Regardless, our observations
would suggest that an individual's resident bacteria have a large role
to play in disease initiation and could change in a way that allows
predictive models to lower the positive probability of a lesion after
surgery {[}Figure 4{]}. It should be noted that our study does not argue
against the importance of these CRC associated bacteria in the
pathogenesis of disease but rather that the models do not utilize these
specific bacteria for classification purposes (lesion or before sample).
In fact, it is possible that these CRC associated bacteria are important
in the transition from adenoma to carcinoma and would be one explanation
as to why in our data we not only see high initial relative abundances,
in certain individuals, but also large decreases in relative abundance
in those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.

>>>>

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or before samples in order to confirm that the community was what was
responding or changing due to surgery. We found that the OTUs that made
up the most important variables to the model overwhelmingly belonged to
comensal bacteria. With only the lesion model having a single OTU from a
previously associated cancer bacterium (\emph{Porphyromonas
asaccharolytica}). Using only these important OTUs and FIT both models
(lesion and before sample) had a significantly decreased positive
probability of either lesion or being a before sample on follow up
{[}Figure 4 \& S4{]}. Further confirmation of the importance of the
changes of comensal bacteria to these classifications was that a total
of 14 OTUs were common to both models and the vast majority belonged to
regular residents of our gut community.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occured. First, one potential explanation is that even with surgery
and a shift of the bacterial community these specific bacteria still
persisted within the colon. Second, the bacteria even if they were
reduced in relative abundance they were still present within the gut.
Third, is that they were not present in enough individuals to be able to
classify those with and without disease with a high degree of accuracy.
Fourth, is that it is possible that our Random Forest models were able
to get the same information from measures such as FIT or other OTUs.
Finally, it is also possible that all of these potential explanations
could have played a role. Regardless, our observations would suggest
that an individual's resident bacteria have a large role to play in
disease initiation and could change in a way that allows predictive
models to lower the positive probability of a lesion after surgery
{[}Figure 4{]}. It should be noted that our study does not argue against
the importance of these CRC associated bacteria in the pathogenesis of
disease but rather that the models do not utilize these specific
bacteria for classification purposes (lesion or before sample). In fact,
it is possible that these CRC associated bacteria are important in the
transition from adenoma to carcinoma and would be one explanation as to
why in our data we not only see high initial relative abundances, in
certain individuals, but also large decreases in relative abundance in
those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.
